---
title: "An additive GP model using lgpr2"
author: "Juho Timonen"
date: 17th September 2024
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{An additive GP model using lgpr2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(lgpr2)
```

# Test data

We load test data.
```{r data}
dat <- testdata
head(dat)
```

# Model definition

We create an additive GP model with many terms.

```{r model}
m <- LonModel$new(y ~ gp(time) + gp(time, arm) + gp(time, id) + gp(time, sex) + gp(weight))
```

# Model fitting

We fit the model.
```{r fit, fig.width=7, fig.height=6}
f <- m$fit(data = dat, chains = 1, iter_warmup = 600, iter_sampling = 600)
f$plot(facet_by = "id")
```

# Visualizing the inferred effects

We visualize individual components to understand effects of the covariates.

```{r plots, fig.width=7, fig.height=3.7}
f$function_draws("f_gp_time")$plot() # shared time effect
f$function_draws("f_gp_timeXarm")$plot() # arm-specific time-effect
f$function_draws("f_gp_timeXsex")$plot() # sex-specific time-effect
```


We see that the effects over different arms and sexes sum to zero at each
time point, due to the zero-sum kernel.

```{r pw, fig.width=6, fig.height=3.7}
f$function_draws("f_gp_weight")$plot()
```


```{r plots2, fig.width=7, fig.height=6}
f$function_draws("f_gp_timeXid")$plot() # individual-specific time-effect
```

The individual-specific effects also sum to zero at each
time point, but it is harder to see it since there are so many subjects.


# Stan code


We print the model Stan code here.
```{r stancode}
cat(m$create_stancode())
```


