---
title: "Mathematical description of lgpr2 models"
author: "Juho Timonen"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Mathematical description of lgpr2 models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(lgpr2)
```


This section describes mathematically the model corresponding to 
the `LonModel` class.

# Observation model

With `GaussianLikelihood`, the likelihood of observation $i$ is
$$
y_i\sim \mathcal{N}(h_i, \sigma^2),
$$
where $h_i = f_i + \hat{C}$, $\hat{C}$ being set to the mean of the 
$y_i$ data and

$$
f_i = f(\mathbf{x}_i)= \sum_{j=1}^J f^{(j)}(\mathbf{x}_i),
$$
$\sigma$ is an unknown parameter, and 
the $\delta$ value is a constant. We use $y_i$ to denote the observed
longitudinal measurement, and $\mathbf{x}_i$ the corresponding covariate
vector. Typically, $\mathbf{x}_i$ includes a subject
identifier $\text{id}_i$, measurement time $t_i$ 
and possible other factors or continuous covariates.
The functions $f^{(j)}$, $j=1, \ldots, J$ are the additive function
components and they are detailed next.

# Model terms


```{r}
create_term <- function(term_str) {
  m <- LonModel$new(as.formula(paste0("y ~ ", term_str)), compile = FALSE)
  m$term_list$terms[[1]]
}
```

The model can have a different number $J$ of terms, depending on the 
`formula` that is given.

## `GroupedOffsetTerm`

```{r}
term <- create_term("offset(g)")
print(term)
```

The `GroupedOffsetTerm` for a grouping variable $g$ is
$$
f^{\text{BAS}} \left(\text{g} \mid \mathbf{c}_{0}\right)
$$
where $\mathbf{c}_{0}$ is an unknown parameter vector containing one
parameter for each level of the grouping variable. These parameters have
independent priors.


## Shared GP terms (`GPTerm`)

```{r}
term <- create_term("gp(x)")
print(term)
```

Let $x$ be some continuous variable. We use
$$
f^{\text{GP}}(x \mid \alpha, \ell) \sim \mathcal{GP}(0,\ k_{\text{EQ}}(x,x' \mid \alpha, \ell)).
$$
to denote that a function is modeled as a zero-mean Gaussian process (GP),
with the exponentiated quadratic kernel function
$$
k_{\text{EQ}}(x, x' \mid \alpha, \ell) = \alpha^2 \exp\left( -\frac{(x-x')^2}{2\ell^2}\right),
$$
where the parameters $\alpha > 0$ and $\ell > 0$ are the kernel magnitude and 
lengthscale parameters, respectively. For GP terms, we use a Hilbert space
basis function approximation
$$
f^{\text{GP}}(x \mid \alpha, \ell) \approx f^{\text{HSGP}}(x \mid \mathbf{\xi}, \alpha, \ell, B, L) = \sum_{b=1}^{B} \xi_b S_{\text{EQ}}\left(\sqrt{\lambda_{L,b}} \mid \alpha, \ell \right) \phi_{L,b}(x),
$$
 where the auxiliary parameters 
$\mathbf{\xi} = \left\{\xi_1, \ldots, \xi_B\right\}$ are a priori standard
normal distributed $\xi_b \sim \mathcal{N}(0,1)$. This approximation approaches
the exact GP in the limit $L \rightarrow \infty$, $B \rightarrow \infty$. The 
basis functions 
$$
\phi_{L,b}(x) = \frac{1}{\sqrt{L}} \sin \left( \frac{\pi b (x + L)}{2L}\right)
$$
are eigenfunctions of a certain Dirichlet boundary value problem,
and $\lambda_{L, b} = \left(\frac{\pi b}{2 L} \right)^2$ are their
corresponding eigenvalues. The function
$$
S_{\text{EQ}}(\omega \mid \alpha, \ell) = \alpha^2 \ell \sqrt{2 \pi} \exp \left(-\frac{\ell^2 \omega^2 }{2}  \right)
$$
is the spectral density of the exponentiated quadratic kernel.The approximation 
domain size $L$ and number of basis functions $B$ are 
constants that are set beforehand. The approximation is accurate only on
the interval $x \in (-L, L)$.

The unknown parameters of a shared GP term are $\alpha, \ell$ and 
$\mathbf{\xi} \in \mathbb{R}^B$.

## Group-specific GP terms (`GPTerm`)

```{r}
term <- create_term("gp(x, z)")
print(term)
```

Let $x$ be some continuous variable and $z$ some grouping factor, which can
without loss of generality take integer values $z \in \{1, \ldots, G\}$. A 
group-specific GP term is
$$
f^{\text{HSGP}}(x, z \mid \mathbf{\xi}_1, \ldots, \mathbf{\xi}_G, \alpha, \ell, B, L) = 
\begin{cases}
f^{\text{HSGP}}(x \mid \mathbf{\xi}_1, \alpha, \ell, B, L) & \text{if } z = 1\\
\vdots \\
f^{\text{HSGP}}(x \mid \mathbf{\xi}_G, \alpha, \ell, B, L) & \text{if } z = G
\end{cases}
$$
which approximates the GP
$$
f^{\text{GP}}(x,z) \sim \mathcal{GP}(0,\ k_{\text{CAT}}(z,z') \cdot k_{\text{EQ}}(x,x')),
$$
where $k_{\text{CAT}}(z,z') = 1$ if $z=z'$ and zero otherwise. The unknown
parameters of a group-specific GP term are $\alpha, \ell$ and a different vector 
$\mathbf{\xi}_g \in \mathbb{R}^B$ for each group $g=1, \ldots, G$.
